Rotation effects on the Lyman-$\alpha$ line morphology in distant galaxies

### 2013-07-03 paper/Analisis.tex
corrected some typos

An ab-initio description of realistic rotation curves in simulations
depends on having access to the dynamic evoluction the mass components
in the galaxy: stars, gas and dark matter. Such level of realism is
extremely complex to achive, specially of one wants to get a
systematic description based on statistics of simulated objects.

Following the tradition of studies of Lyman$\alpha$ emitting systems,
we implement a model with a simplified geoemtry and gas
distribution. We assume that the gas is homogeneously distributed in a
sphere that rotates as a solid body with constant angular
velocity. This simple model will contain only one parameter: the
linear velocity at the sphere's surface, $V_{\rm max}$.

@@

 In the MonteCarlo code we define a cartesian coordinate system to
 define the position of each photon. The origin of this system
 coincides withthe center of the sphere and the rotation axis is defined
 to be $z$-axis. With this choice, the components of the gas bulk velocity
 field, $\vec{v} = v_{x}\hat{i} + v_{y}\hat{j} + v_{z}\hat{k}$, can be
 written as

@@

In contrast with spherically symmetric models (static, outflow,
inflow) the rotation now defines a preferred direction in the
problem. In Section \ref{sec:results} we quantify this differences by
varying the line of sight of a mock observer with respect to the
rotation axis. The results are parameterized by the polar angle
$\theta$ as defined by the dot product $\cos\theta =
{\hat{u}\cdot\hat{k}}$.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies, each one
varying in at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, hydrogen optical depth $\tau_{H}$,
dust optical depth $\tau_{a}$ and the initial disitrbution of photons
with respect to the gas. There are in total 60 models with the input
parameters summarized in Table \ref{table:models}.

@@

The central result of this paper is summarized in
Fig.~\ref{fig:differentvelocities} that shows clearly the considerable
impact of rotation on the morphology of the emergent \ly line. Both
panels in the Figure focus on the results for $\tau_{H}=10^{7}$,
showing that the influence of rotation is present both when the
photons are either homogeneusly or centrally initialized over the gas
volume.

In the following subsections we characterize the line morphology by
the half-width at half intensity and the peak maxima. In order to
intepret the morphological changes in the line we also report the
median number of scatter for each \ly photon in the
simulation. Finally we measure the bulk escape fraction as a function
of rotational velocity in the presence of dust.

@@

\begin{figure}
    \includegraphics[width=0.45\textwidth]{WidthvsVmax.png}
    \caption{Half-width for the non-dusty models as a function of
      rotational velocity $V_{\rm max}$. Continous (dashed) lines
      correspond to homogeneous (central) source
      distributions. \label{fig:widthvsvelocity}} 
\end{figure}

\begin{figure}
    \includegraphics[width=0.45\textwidth]{maximumvsVmax.png}
\caption{Position of the peak maxima as a function of rotational
  velocity $V_{\rm max}$. Continous (dashed) lines correspond to
  homogeneous (central) source distributions. A value of $x_{\rm
    max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumsvsvelocity}}  
\end{figure}

The first quantitative conclusion of the effect of rotation in the
\ly line is that double peaks broaden and reduce their intensity
while the line center rises. This produces the impresion that, as the
rotational velocity increases, the double peaks are merged into a
single broad emission peak. This is msot evident in the highest
rotational velocities for the homogeneously distributed sources
(right panel in Fig.~\ref{fig:differentvelocities}).

@@

Figure \ref{fig:widthvsvelocity} shows how $W_{1/2}$ increases with
rotational velocity. Continous (dashed) lines connect the results for
homogeneous (central) source distribution. The line width increases
with rotational velocity up to factors of $\sim 2$ at $300$\kms with
respect to the static case. For the temperature $T=10^4$K used in our
radiative transfer calculations calculations the thermal velocity is
$v_{th}=12.8$\kms. For a model with $\tau_{H}$ it means that the
half-width can increase up to $350$\kms (at $V_{\rm max}=300$\kms)
compared to a half-width of $150$\kms in the static case.

@@

For larger values of $\tau_{H}$ the transitional velocity $V_{\rm
  trans}$ is also larger. In our models we find the following
correspondance between the optical depth $\tau_{\rm H}=\{10^5, 10^6,
10^7\}$ and the transitional velocities $V_{\rm
  trans}=\{50-100\kms, 100-200\kms, 200-300\kms\}$ which can only be
constrained to be in the range of velocities in the models that gave a $x_{\rm
  max}\neq 0$ and $x_{\rm max}=0$.   In the next section explore the
origin of this trends and consider to what extent the hypothesis of a low
number of scatterings is correlated with the emergence of a single
peak.

@@

The number of times that a \ly photon is absorbed and re-emmited is
connected to the final frequency that it can have after escaping the
galaxy. In the case of static gas geometries, a large value of the
optical depth is immediatly followed by a high number of
scatterings. In turn a large optical depth increases the probability
that a \lya photon will be found far from the center of the line. The
peak maxima shifting away form the line center as the amount of
neutral hydrogen increases.

In Figure \ref{fig:Nscatt} we see the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central distributions we find that there is not
a significante change for increasing rotational
velocities, $\langle N_{\rm scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this case we also find that the average
number of scatterings is proportional to the optical depth, as
expected in analogy from the analytic result for the homogeneous
infinite-slab $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}. In our experiments we find  $\langle N_{\rm
  scatt}\rangle= XXX\tau_{\rm   H}$.

@@

In order to gain a deeper understanding of these results we prepare 2D
histograms for the number of scaterrings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
the results of such histogram in the case $\tau{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$. This Figure clearly support our hypothesis about
the photosphere in the homogeneous distribution, in this case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings, clearly explaining the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, on the
order of the optical depth, but the two peaks do get closer to each
other. In this situation each scattering leaves the photon closer to
the center of the line than the static case.

@@

\begin{figure}
  \includegraphics[width=0.45\textwidth]{escapefraction.png}
   \caption{Escape fraction as a function of rotational velcity. The
     continous (dashed) lines correspond to homogeneous (central)
     models.
     \label{fig:efvsv}}
\end{figure}

\begin{figure}
  \includegraphics[width=0.45\textwidth]{Neufeld.png}
 \label{fig:efvsNeufeld}\caption{Escape fraction as a function of the
   product $(a\tau_{\rm H})^{1/3}\tau_{a}$. The analytic solution for
   the infinite is slab shown as a continous line. Different dashed
   lines correspond to different rotational velocities.}   
\end{figure}

We also study a dusty cloud configuration to measure the effect of
rotation on the escape fraction. We expect that the modified number of
scatterings to be reflected in this amount of photons aborbed by
dust. Following this line of thought,  we do not expect any change in
for a dusty cloud with central source of radiation given that the
number of scatterings remains close to constant. On the other hand, in
the case of an homogenous radiation source the number of scatterings
drops as $V_{\rm max}$ increases, which might be reflected as an
increasing escape farction.

@@

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytical solution for the infinite slab \citep{Neufeld90}. In
Neufeld's setup the analytic solution depends solely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} corespond to the cases of different
velocities. We observe that the escape fraction is higher
by factors of $2-10$  than the expected values for the slab
configuration. We also see that for the lower value $\tau_{\rm
  H}=10^{5}$ the escape fraction does not increase with respect to the
solution for $\tau_{\rm H}=10^{6}$ as expected, however we not that we
are in a regime where the condition for the analytic expectations 
($a\tau_{h}\gg 1$) does not hold.

>>>>

An ab-initio description of realistic rotation curves in simulations
depends on having access to the dynamic evolution the mass components
in the galaxy: stars, gas and dark matter. Such level of realism is
extremely complex to achieve, specially of one wants to get a
systematic description based on statistics of simulated objects.

Following the tradition of studies of Lyman$\alpha$ emitting systems,
we implement a model with a simplified geometry and gas
distribution. We assume that the gas is homogeneously distributed in a
sphere that rotates as a solid body with constant angular
velocity. This simple model will contain only one parameter: the
linear velocity at the sphere's surface, $V_{\rm max}$.

@@

 In the MonteCarlo code we define a cartesian coordinate system to
 define the position of each photon. The origin of this system
 coincides with the center of the sphere and the rotation axis is defined
 to be $z$-axis. With this choice, the components of the gas bulk velocity
 field, $\vec{v} = v_{x}\hat{i} + v_{y}\hat{j} + v_{z}\hat{k}$, can be
 written as

@@

In contrast with spherically symmetric models (static, outflow,
inflow) the rotation now defines a preferred direction in the
problem. In Section \ref{sec:results} we quantify this differences by
varying the line of sight of a mock observer with respect to the
rotation axis. The results are parametrized by the polar angle
$\theta$ as defined by the dot product $\cos\theta =
{\hat{u}\cdot\hat{k}}$.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies, each one
varying in at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, hydrogen optical depth $\tau_{H}$,
dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. There are in total 60 models with the input
parameters summarized in Table \ref{table:models}.

@@

The central result of this paper is summarized in
Fig.~\ref{fig:differentvelocities} that shows clearly the considerable
impact of rotation on the morphology of the emergent \ly line. Both
panels in the Figure focus on the results for $\tau_{H}=10^{7}$,
showing that the influence of rotation is present both when the
photons are either homogeneously or centrally initialized over the gas
volume.

In the following subsections we characterize the line morphology by
the half-width at half intensity and the peak maxima. In order to
interpret the morphological changes in the line we also report the
median number of scatter for each \ly photon in the
simulation. Finally we measure the bulk escape fraction as a function
of rotational velocity in the presence of dust.

@@

\begin{figure}
    \includegraphics[width=0.45\textwidth]{WidthvsVmax.png}
    \caption{Half-width for the non-dusty models as a function of
      rotational velocity $V_{\rm max}$. Continuous (dashed) lines
      correspond to homogeneous (central) source
      distributions. \label{fig:widthvsvelocity}} 
\end{figure}

\begin{figure}
    \includegraphics[width=0.45\textwidth]{maximumvsVmax.png}
\caption{Position of the peak maxima as a function of rotational
  velocity $V_{\rm max}$. Continuous (dashed) lines correspond to
  homogeneous (central) source distributions. A value of $x_{\rm
    max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumsvsvelocity}}  
\end{figure}

The first quantitative conclusion of the effect of rotation in the
\ly line is that double peaks broaden and reduce their intensity
while the line center rises. This produces the impression that, as the
rotational velocity increases, the double peaks are merged into a
single broad emission peak. This is most evident in the highest
rotational velocities for the homogeneously distributed sources
(right panel in Fig.~\ref{fig:differentvelocities}).

@@

Figure \ref{fig:widthvsvelocity} shows how $W_{1/2}$ increases with
rotational velocity. Continuous (dashed) lines connect the results for
homogeneous (central) source distribution. The line width increases
with rotational velocity up to factors of $\sim 2$ at $300$\kms with
respect to the static case. For the temperature $T=10^4$K used in our
radiative transfer calculations calculations the thermal velocity is
$v_{th}=12.8$\kms. For a model with $\tau_{H}$ it means that the
half-width can increase up to $350$\kms (at $V_{\rm max}=300$\kms)
compared to a half-width of $150$\kms in the static case.

@@

For larger values of $\tau_{H}$ the transitional velocity $V_{\rm
  trans}$ is also larger. In our models we find the following
correspondence between the optical depth $\tau_{\rm H}=\{10^5, 10^6,
10^7\}$ and the transitional velocities $V_{\rm
  trans}=\{50-100\kms, 100-200\kms, 200-300\kms\}$ which can only be
constrained to be in the range of velocities in the models that gave a $x_{\rm
  max}\neq 0$ and $x_{\rm max}=0$.   In the next section explore the
origin of this trends and consider to what extent the hypothesis of a low
number of scatterings is correlated with the emergence of a single
peak.

@@

The number of times that a \ly photon is absorbed and re-emitted is
connected to the final frequency that it can have after escaping the
galaxy. In the case of static gas geometries, a large value of the
optical depth is immediately followed by a high number of
scatterings. In turn a large optical depth increases the probability
that a \lya photon will be found far from the center of the line. The
peak maxima shifting away form the line center as the amount of
neutral hydrogen increases.

In Figure \ref{fig:Nscatt} we see the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central distributions we find that there is not
a significant change for increasing rotational
velocities, $\langle N_{\rm scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this case we also find that the average
number of scatterings is proportional to the optical depth, as
expected in analogy from the analytic result for the homogeneous
infinite-slab $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}. In our experiments we find  $\langle N_{\rm
  scatt}\rangle= XXX\tau_{\rm   H}$.

@@

In order to gain a deeper understanding of these results we prepare 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
the results of such histogram in the case $\tau{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$. This Figure clearly support our hypothesis about
the photosphere in the homogeneous distribution, in this case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings, clearly explaining the origin of a single central
peak. However, for a central distribution the situation is
different. In this case the number of scatterings remains high, on the
order of the optical depth, but the two peaks do get closer to each
other. In this situation each scattering leaves the photon closer to
the center of the line than the static case.

@@

\begin{figure}
  \includegraphics[width=0.45\textwidth]{escapefraction.png}
   \caption{Escape fraction as a function of rotational velocity. The
     continuous (dashed) lines correspond to homogeneous (central)
     models.
     \label{fig:efvsv}}
\end{figure}

\begin{figure}
  \includegraphics[width=0.45\textwidth]{Neufeld.png}
 \label{fig:efvsNeufeld}\caption{Escape fraction as a function of the
   product $(a\tau_{\rm H})^{1/3}\tau_{a}$. The analytic solution for
   the infinite is slab shown as a continuous line. Different dashed
   lines correspond to different rotational velocities.}   
\end{figure}

We also study a dusty cloud configuration to measure the effect of
rotation on the escape fraction. We expect that the modified number of
scatterings to be reflected in this amount of photons absorbed by
dust. Following this line of thought,  we do not expect any change in
for a dusty cloud with central source of radiation given that the
number of scatterings remains close to constant. On the other hand, in
the case of an homogeneous radiation source the number of scatterings
drops as $V_{\rm max}$ increases, which might be reflected as an
increasing escape fraction.

@@

In Figure \ref{fig:efvsNeufeld} we put these results in the context of
the analytical solution for the infinite slab \citep{Neufeld90}. In
Neufeld's setup the analytic solution depends solely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{a}$, an approximation that is valid only
in the limits $a\tau_{\rm   H}\gg 1$. The dashed lines in Figure
\ref{fig:efvsNeufeld} correspond to the cases of different
velocities. We observe that the escape fraction is higher
by factors of $2-10$  than the expected values for the slab
configuration. We also see that for the lower value $\tau_{\rm
  H}=10^{5}$ the escape fraction does not increase with respect to the
solution for $\tau_{\rm H}=10^{6}$ as expected, however we not that we
are in a regime where the condition for the analytic expectations 
($a\tau_{h}\gg 1$) does not hold.
